---
title: "BaRcodes - tracking and labelling biological samples"
author: "Yihan Wu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start to Use BaRcodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

introduction points:
biological samples are complex to label and maintain.
Multi-level sampling, ie. populations, lines, individuals, samples per individual.
Manual creation of barcodes, especially hierarchical ones, are timeconsuming. Commercial software requires extra installation and is difficult to customize.  



The BaRcodes package will create barcodes in a pdf output to be printed using a laser printer and sticker sheets. Each function can be run interactively with prompts to guide label and barcode creation. 


If you have text labels in a one column text file, skip to {Create barcodes}.

Required packages 

'BaRcodes' requires four additional packages and their dependencies. 

```{r}
install.packages(c("ggplot2", "grDevices", "grid", "qrcode"))
```


Graphic of simple vs hierarchical labels?

Create simple labels  

Simple one-level barcodes can be created using the 'label_maker()' function. You can use the function interactively.
```{r}
Labels<-label_maker(user=T)
```

User prompts will be shown in the console. 

Create hierarchical labels

Create barcodes

Advanced Options

