---
title: "BaRcodes - tracking and labelling biological samples"
author: "Yihan Wu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start to Use BaRcodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

introduction points:
biological samples are complex to label and maintain.
Multi-level sampling, ie. populations, lines, individuals, samples per individual.
Manual creation of barcodes, especially hierarchical ones, are timeconsuming. Commercial software requires extra installation and is difficult to customize.  



The BaRcodes package will create barcodes in a pdf output to be printed using a laser printer and sticker sheets. Each function can be run interactively with prompts to guide label and barcode creation. 


If you have text labels in a one column text file, skip to {Create barcodes}.

Required packages 

'BaRcodes' requires four additional packages and their dependencies. 

```{r eval=F, include=T}
install.packages(c("ggplot2", "grDevices", "grid", "qrcode"))
```


Graphic of simple vs hierarchical labels?

# Create simple labels {#label-maker}

Simple one-level barcodes can be created using the 'label_maker()' function. One-level labels consist of two parts, a string and a number. The string will remain the same for all labels produced while the number will vary. This can be used to label and track samples from a population where each individual will be assigned a different number.

The purpose of the example below will be to create labels like below. 

```
example005
example006
example007
example008
example009
example010
```
First, load `BaRcodes`.

```{r eval=F, include=T}
library(BaRcodes)
```

## By user prompt {#label-maker-user-prompt}

```{r eval=F, include=T}
Labels<-label_maker(user=T)
```

In the code snippet above, user prompts will be shown in the console and the user input to the prompts will set the parameters to create labels that will be contained in the object `Labels`. The first user prompt shown in the console will be: 

>  \> Please enter string for level: 

The string will be the part of the label that will be the same for every label produced. This can be the name of a population, a species or a location. In this example, the string entered by the user into the console as denoted by the `>` symbol is "example."

>  \> Please enter string for level: example

The second user prompt is:

>  \> Enter the starting number for level: 

The prompt asks for the smallest number for the labels. The response entered was 5. The user has the option of determining which numbers the label should start increasing from. Labels produced do not have to start from 1. 

>  \> Enter the starting number for level: 5

The third user prompt is:

>  \> Enter the ending number for level: 

The prompt asks for the largest number suffix. The labels produced will be numbered sequentially from the starting number to the ending number. It is possible to create labels that are not sequentially numbered (see [By arguments](#label-maker-arguments)) Here, the user input is 10. 

>  \> Enter the ending number for level: 10

After the starting and ending numbers are set, the function calculates the number of digits in the biggest number for the labels. In this example, the maximum number of digits is 2. The console will print:

> Please enter a number larger or equal to 2

and ask:

>  \> Number of digits to print for level:

It is possible to set the number of digits to print for the labels. This may come in useful for consistency if similar labels with higher numbers is produced in the future. In this example, the number of digits printed is set to 3.

>  \> Number of digits to print for level: 3 

A string of characters will be returned. 

## By arguments {#label-maker-arguments}
It is also possible to create the labels directly without user prompts by the arguments in `label_maker`. The example can also be reproduced by the code:

```{r eval=F, include=T}
Labels<-label_maker(string="example", level=5:10, digits=3)
```

Because the number can be set through the argument 'level', it is possible to use a non-sequential sequence of numbers such as below:

```{r eval=F, include=T}
number_sequence<-seq(1, 10, 2)
Labels<-label_maker(string="example", level=number_sequence, digits=3)
```

The output will then be:

> [1] "example001" "example003" "example005" "example007" "example009"

Labels can then be saved using:

```{r eval=F, include=T}
write.csv(Labels, "Labels.csv")
```


# Create hierarchical labels {#label_hier_maker}

`label_hier_maker` is used to make labels that follow a hierarchical or tree-like structure. Similar to `label_maker`, this function can be run interactively and also using the arguments in the function. It is possibel to set different numbers of levels, each with their own text string that is consistent across the level, and individual numerical suffixes. Below is an example of a list of hierarchical labels with three levels (a, b, c) and varying numbers of individuals for each level (a=3, b=2, c=2). 

```
a1-b1-c1
a1-b1-c2
a1-b2-c1
a1-b2-c2
a2-b1-c1
a2-b1-c2
a2-b2-c1
a2-b2-c2
a3-b1-c1
a3-b1-c2
a3-b2-c1
a3-b2-c2
```



## By user prompts {#label-hier-maker-user-prompt}

To create the hierarchical labels using prompts, start with the argument `user=T` in the function.

```{r eval=F, include=T}
Labels<-label_hier_maker(user=T)
```

The prompt that appears in the console is:

>  \> What is the # of levels in hierarchy: 

In this example, we have levels a, b, and c, three levels in total. 

>  \> What is the # of levels in hierarchy: 3

The second prompt asks if a string should be appended to all labels.

>  \> String at end of barcode? (y/n):

There are only two possible responses to this question as indiciated in the parenthesis. If the answer is yes, then a new prompt will appear.

>  \> Please enter ending string: 

In this example, there is no ending string.

>  \> String at end of barcode? (y/n): n

For each level, there will be a series of prompts. The user will be able to set the string, the starting number, and the ending number for each level similar to the questions in `label_maker`(see [label_maker](#label-maker-user-prompt) for step by step instructions). The number of digits to print for each level can also vary. 

## By argument

It is also possible to specify arguments directly into `label_hier_maker`. Unlike `label_maker`, a list of the hierarchy parameters has to be created first. Vectors should be created for each level in the hierarchy. Each vector will contain three values, the string, the beginning value, and the ending value. The vectors should be concatenated together based on their order. 

```{r}
level_one<-c("a", 1, 3)
level_two<-c("b", 1, 2)
level_three<-c("c", 1, 2)
hier_list<-list(level_one, level_two, level_three)
```

You can specify the ending string for all labels through the `end` argument and the number of digits to print for all levels through the `digits` argument. It is not possible at this time to vary the number of digits printed for each level, something that is possible when using the prompts. 

The list can then be passed into `label_hier_maker`.

```{r, eval=F, include=T}
Labels<-label_hier_maker(hierarchy = hier_list, digits=1)
```

The output will be:

> [1] "a1-b1-c1" "a1-b1-c2" "a1-b2-c1" "a1-b2-c2" "a2-b1-c1" "a2-b1-c2" "a2-b2-c1" "a2-b2-c2" "a3-b1-c1" "a3-b1-c2" "a3-b2-c1" "a3-b2-c2"

Labels can then be saved using:

```{r eval=F, include=T}
write.csv(Labels, "Labels.csv")
```



# Create barcodes {#create-barcodes}

2D barcodes are created using a set of text labels. Users can manually create their own labels in a one-column format or create labels using `label_maker` or `label_hier_maker`. 

The function `create_PDF` has been set up to produce a pdf file containing barcodes that fit to the dimensions of ULINE Premium Laser Printer Labels #S-19297 (80 labels per sheet, 1.75" x 0.5"). If needed, the page setup can be modified using advanced options in [custom_create_PDF](#custom-create-pdf). 

The first step is to read in the labels from file:

```{r eval=F, include=T}
Labels<-read.csv("Labels.csv")
```

## By user prompt {#create-barcodes-user-prompt}

Here, using the hierarchical set of labels, a PDF file containing the labels will be created with the name "example.pdf", a font size of 3.5, and an error correction level "Q" meaning the barcode can tolerate up to 25% damage. 
with `user=T`, user promps will show up to guide creation of a PDF file containing barcodes.

```{r eval=F, include=T}
create_PDF(user=T)
```
A user prompt is printed into the console. For example:

>  \> Please enter name for PDF output file: 

Any combination of letters and numbers can be used for the name of the pdf file. Here, the file name is set to "example."

>  \> Please enter name for PDF output file: example

The next user prompt is to set the size of the text printed on each barcode. 

>  \> Please enter a font size (2.2-4.7): 

This user input has to be a number between 2.2 and 4.7. In this example, the font size will be 3.5.

>  \> Please enter a font size (2.2-4.7): 3.5

It is possble that the labels are too long, or cannot fit on one row due to the size of the font selected. This will cause an error to print to console.

> ERROR: not enought space to print full label, please decrease font size.

The font size user prompt will appear again to reset to a more appropriate font size.  

>  \> Please enter a font size (2.2-4.7):

The last basic parameter to set is the error correction level. There are four possible levels: L, M, Q, and H. 

Level "L" - up to 7% damage -- ideal for very small labels (large pixels)
Level "M" - up to 15% damage
Level "Q" - up to 25% damage
Level "H" - up to 30% damage -- good for bigger levels (small pixels)

The user prompt for error correction level is similar to previous prompts.

>  \> Specify an error correction - L, M, Q, H: 

This example uses an error correction level "Q"

>  \> Specify an error correction - L, M, Q, H: Q

The last user prompt asks whether the user wants to modify the advanced parameters. 

>  \> Edit advanced parameters? (Y/N): 

In this example, no advanced parameters are modified. The advanced parameters are covered in [advanced options](#custom-create-pdf)

>  \> Edit advanced parameters? (Y/N): N

# Advanced Options of for pdf output {#custom-create-pdf}

